Analyses

### 2023-05-12 manuscript_with-deletions.RmD
- major updates before submission - language, analyses, commenting

abstract: |
    The privacy calculus model states that people weigh risks and benefits before communicating online. This paper analyzes three open questions. How robust is the privacy calculus, can it be replicated with actual behavior in realistic online setting? Do we need to extend the model by integrating more socio-psychological measures such as trust, self-efficacy, and privacy deliberation? How easily can the privacy calculus be be manipulated by external factors such as like and dislike buttons? In a preregistered one-week field experiment (N = 590), participants discussed a current political topic on an online website, which included either (a) like buttons, (b) both like and dislike buttons, or (c) no buttons. The privacy calculus model was confirmed: Benefits and concerns affected communication. Deliberating about privacy by comparing benefits and risks decreased communication. Experiencing self-efficacy and trust increased communication. Implementing both like and dislike buttons had a strong effect on behavior: It reduced online communication by 45 percent.

@@

bibliography      : 
  - [bibliography/r-references.bib]
  - [bibliography/references.bib]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
draft             : no
documentclass     : "apa6"
classoption       : "man"
# output            : papaja::apa6_pdf
# header-includes:
#   - \setlength{\parskip}{0em}
#   - \raggedbottom
#   - \note{\clearpage}
# output            : papaja::apa6_docx
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
---

@@

According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
People communicate much personal information online, although being strongly concerned about their privacy [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@dienlinPrivacyParadoxAus2019] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is tenuous [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021]. 
Research is increasingly building on the privacy calculus model, which states that communication online can be explained by perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018b; @krasnovaOnlineSocialNetworks2010; @meierPrivacyCalculusRevisited2022].

@@

Second, the privacy calculus model is criticized for neglecting the actual process of weighing pros and cons, and for over-emphasizing rationality  [@knijnenburgDeathPrivacyCalculus2017].
Simply showing that both concerns and gratifications correlate with communication online does not prove that an explicit weighing process took place.
Instead, we need to find out if people deliberately compare costs and benefits before communicating online. 
Therefore, I will also analyze the process of comparing pros and cons explicitly, by elaborating on what I will call the the privacy deliberation process [@omarzuDisclosureDecisionModel2000]. 
Next, focusing exclusively on costs and benefits does not fully capture how and why users engage online.
I, hence, extend the privacy calculus by analyzing how online communication is affected by trust and self-efficacy, two variables less focused on rationality and more oriented toward socio-psychological aspects [@metzgerPrivacyTrustDisclosure2004].

Online behavior is determined by the person itself and their external circumstances [@bazarovaIntegrationIndividualisticNetworked2020; @treptePrivacyCalculusContextualized2020; @spottswoodShouldShareThat2017].
The privacy calculus upholds that much of behavior is rational and self-determined, but how easily can this process be manipulated by external circumstances?
Online contexts are optimized to elicit as much interaction as possible [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021], for example via low threshold communication features such as likes, shares, replies, or reactions  [@carrPredictingThresholdPerceived2018]. 
Implicit and explicit cues on how a website is used can increase self-disclosure [@spottswoodShouldShareThat2017; @treptePrivacyCalculusContextualized2020].
How large are the effects of external popularity cues such as like and dislike buttons compared to the internal weighing of pros and cons?
And how easily can the privacy calculus be manipulated by popularity cues?

@@

The privacy calculus analyzes why people communicate online.
When are we willing to engage in a conversation?
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977] and states that people are weighing risks and benefits before engaging in communication.
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Disclosing information can be risky, and if people are more concerned about their privacy 
they are less likely to communicate online [@baruhOnlinePrivacyConcerns2017; @krasnovaOnlineSocialNetworks2010; @masurUnderstandingEffectsConceptual2023; @meierPrivacyCalculusRevisited2022].
According to a meta-analysis by @baruhOnlinePrivacyConcerns2017, the effect is small but significant (*r* = -.13).

Outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016a; @bolUnderstandingEffectsPersonalization2018b; @kezerGettingPrivacyCalculus2022].
People are happy to trade in parts of their privacy when they can receive something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support [@krasnovaOnlineSocialNetworks2010], social capital [@ellisonBenefitsFacebookFriends2007], entertainment [@dhirUnderstandingRelationshipIntensity2017], information-seeking [@whitingWhyPeopleUse2013], and self-presentation [@minHowArePeople2015].

@@

Although privacy calculus implies that people explicitly compare benefits and risks before communicating online, prior research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if, and if so by how much, people actively deliberate about their privacy, and whether doing so influences their decision to communicate.
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits of communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.

How could deliberating about one's privacy affect communication?
On the one hand, it could reduce subsequent communication.
Refraining from communication---the primary means of connecting with others [@altmanPrivacyConceptualAnalysis1976]---often requires restraint [@omarzuDisclosureDecisionModel2000]. 
This is especially true for social media, which are designed to foster communication and participation [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021].
Actively thinking about whether communicating is worthwhile might be the first step not to participate.
On the other hand, deliberating about privacy might also increase communication.
In light of the numerous benefits mentioned above, it might make sense to conclude that participation is beneficial, thereby fostering communication [@krasnovaOnlineSocialNetworks2010].

@@

The aforementioned weighing process is not flawless or perfect.
It is useful to understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
Importantly, although human behavior is _partly_ irrational, bounded rationality does not state that it is _completely_ irrational [@gigerenzerBoundedRationalityAdaptive2002].
Rationality is a continuum. 
In the context of online privacy, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors next to gratifications and concerns should also explain communication.
Factors less focused on rationality, but more on socio-psychological aspects.

Prior research suggests that perhaps the following two factors are most important when it comes to understanding online communication: self-efficacy and trust [@correaParticipationDivideOnline2010; @hossainICTEnabledVirtual2004; @metzgerPrivacyTrustDisclosure2004].
For example, if users are more familiar, experienced, and knowledgeable in a given online context, they are more likely to navigate that online contexts successfully and to communicate actively [@baruhOnlinePrivacyConcerns2017; @kramerMasteringChallengeBalancing2020].
People with more privacy self-efficacy also engage more successfully in self-withdrawal [@dienlinExtendedPrivacyCalculus2016a].
Hence, if users possess more self-efficacy to participate, they should also communicate more.

@@

In all situations where people lack experience, control, or competence, a central variable to understand behavior is trust [@gefenTrustTAMOnline2003].
Especially in online contexts, trust plays a key role [@metzgerPrivacyTrustDisclosure2004].
Users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
In online contexts, there are different targets of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011].

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
How easily can the calculus be manipulated externally?
One of the central tools to afford and govern online behavior are popularity cues such as like and dislike buttons [@stsiampkouskayaNotExperimentalStudy2023].
Popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020]:
Online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].

How and why might popularity cues affect the privacy calculus?
In analyzing this question, it makes sense to analyze the cues' underlying affordances [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
There is an ongoing debate on how to best understand and operationalize affordances [@evansExplicatingAffordancesConceptual2017].
Whereas @evansExplicatingAffordancesConceptual2017 propose three affordances for mediated communication (i.e., anonymity, persistence, and visibility), @foxDistinguishingTechnologiesSocial2017 suggest 10 affordances for SNSs alone (i.e., accessibility, bandwidth, social presence, privacy, network association, personalization, persistence, editability, conversation control, and anonymity).
Popularity cues such as like and dislike buttons, which are of primary interest in this study, are understood as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142].

Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might afford active communication. 
The mere option to receive dislikes, conversely, might inhibit communication.
Second, _actually receiving_ likes or dislikes might affect behavior, by means of positive reinforcement (likes) or punishment (dislikes) [@skinnerScienceHumanBehavior2014].
To illustrate, likes are affirmative and embody the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like-buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate. 
Dislikes, instead, represent a punishment, introducing a _loss frame_. 
Websites featuring both like _and_ dislike buttons should therefore be more ambivalent compared to websites without popularity cues.
Privacy concerns should not be reduced anymore:
People who are more concerned about their privacy are also more shy and risk averse [@dienlinPsychologyPrivacyAnalyzing2017].
Implementing the dislike button might therefore increase privacy concerns, canceling out the positive effects of the like button.

@@

@@

This manuscript features a [companion website](https://XMtRa.github.io/like_dislike), which includes the data, research materials, analysis scripts, and a reproducible version of this manuscript (see https://XMtRa.github.io/like_dislike).
The hypotheses, sample size, research materials, analyses, and exclusion criteria were preregistered (see https://osf.io/a6tzc/?view_only=5d0ef9fe5e1745878cd1b19273cdf859). 
In some cases, the preregistered approach had to be changed (see [companion website](https://XMtRa.github.io/like_dislike/preregistration_changes)). 
Analyses not preregistered are reported as exploratory analyses.

@@

The data were collected in Germany.
Participants were recruited using the professional panel agency Norstat.
As incentive, participants were awarded digital points, to receive special offers from online retailers. 
Participants had to be above the age of 18 and had to live in Germany. 
In a first step, the company sent its panel members an invitation to participate in the study. 
In this invitation, panel members were asked to participate in a study analyzing the current threat posed by terrorist attacks in Germany.
Members who agreed to participate were sent the first questionnaire (_T1_).
The questionnaire asked participants about their sociodemographic background, (b) provided more details about the study, and (c) included a registration link for the website, which was introduced as "participation platform".
Afterward, participants were randomly assigned to one of the three websites. 
After registration was completed, participants were invited (but not obliged) to visit the website and to discuss the topic of the terrorism threat in Germany. 
Participants could use the website and write comments over the course of one week.
Subsequently, participants received a follow-up questionnaire in which the self-reported measures reported below were collected (_T2_).

@@

Two types of trust were measured.
_General trust_ was operationalized based on @sollnerWhyDifferentTrust2016, addressing three targets (i.e., provider, website, and other users), measured with one item each. 
One example item was "The operators of the participation platform seemed trustworthy."
_Specific trust_ was operationalized for the same three targets with three sub-dimensions each (i.e., ability, benevolence/integrity, and reliability), measured with one item each. 
Example items were "The operators of the participation platform have done a good job" (ability), "The other users had good intentions" (benevolence/integrity), "The website worked well" (reliability). 
Participants placed a lot of trust in the website, the providers and the other users (trust general: _m_ = `r facval_tab["General trust", "m"]`; trust specific: _m_ = `r facval_tab["Specific trust", "m"]`).

>>>>

abstract: |
    The privacy calculus model states that people weigh risks and benefits before communicating online. This paper analyzes three open questions. How robust is the privacy calculus, can it be confirmed with actual behavior in realistic online setting? Do we need to extend the model by integrating more socio-psychological measures including trust, self-efficacy, and privacy deliberation? Can the privacy calculus be be manipulated by external factors such as like and dislike buttons? A preregistered one-week field experiment with 590 participants was conducted. Participants discussed a current political topic on an online website, which included either (a) like buttons, (b) both like and dislike buttons, or (c) no like or dislike buttons. The privacy calculus model was confirmed: Benefits and concerns affected communication. Deliberating about privacy by comparing benefits and drawbacks decreased communication, while experiencing self-efficacy and trust increased communication. Implementing both like and dislike buttons reduced online communication by 44%.

@@

bibliography      : 
  - [bibliography/r-references.bib]
  - [bibliography/references.bib]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
  - \note{\clearpage}
# output            : papaja::apa6_docx
# output:
#   html_document:
#     toc: TRUE
#     toc_float:
#       collapsed: FALSE
#     toc_depth: 2
---

@@

<!-- Understanding why people share personal information online is important.  -->
According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
<!-- Personal beliefs, privacy concerns, or behavioral attitudes do not affect online behavior. -->
<!-- [for an overview of the privacy paradox, see @dienlinPrivacyParadoxAus2019].  -->
People communicate vast amounts of personal information online, although they are concerned about their privacy [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@dienlinPrivacyParadoxAus2019] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is weak [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021]. 
<!-- , which shows that privacy concerns are indeed related to communication online.  -->
<!-- Rather than further pursuing the privacy paradox,  -->
Research is therefore increasingly building on the privacy calculus model, which states that communication online can be explained by perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018b; @krasnovaOnlineSocialNetworks2010; @meierPrivacyCalculusRevisited2022].
<!-- Several studies have shown that experiencing more privacy risks or concerns is related to sharing less information online [@bolUnderstandingEffectsPersonalization2018b; @meierPrivacyCalculusRevisited2022], while expecting more benefits is related to sharing more information online [@heirmanPredictingAdolescentsDisclosure2013; @koohikamaliInvestigationDynamicModel2019]. -->

@@

<!-- I therefore analyze whether the privacy calculus is affected by the affordances of a website.  -->
<!-- Specifically, I investigate whether *popularity cues* such as like and dislike buttons affect the privacy calculus and whether they foster communication online. -->
Second, the privacy calculus model is criticized for neglecting the actual process of weighing pros and for over-emphasizing rationality  [@knijnenburgDeathPrivacyCalculus2017].
Simply showing that both concerns and gratifications correlate with communication online does not prove that an explicit weighing process took place.
Instead, we need to find out if people deliberately compared costs and benefits before communicating online. 
Therefore, I will also analyze the explicit process of comparing pros and cons, by elaborating on what I will call the the privacy deliberation process. 
Next, focusing exclusively on costs and benefits does not sufficiently represent how and why users engage online.
Hence, I extend the privacy calculus in analyzing how online communication is affected by trust and self-efficacy, two variables less focused on rationality and more on socio-psychological aspects [@metzgerPrivacyTrustDisclosure2004].

Online behavior is determined by the person itself and their external circumstances [@bazarovaIntegrationIndividualisticNetworked2020; @treptePrivacyCalculusContextualized2020; @spottswoodShouldShareThat2017].
The privacy calculus upholds that much of behavior is rational and self-determined, but how easily can this process be manipulated by external circumstances?
Online contexts are optimized to elicit as much interaction as possible, for example via low threshold communication features such as likes, shares, replies, or reactions  [@carrPredictingThresholdPerceived2018]. 
Implicit and explicit cues on how others use a website can increase self-disclosure [@spottswoodShouldShareThat2017; @treptePrivacyCalculusContextualized2020].
<!-- The third and last question of this study is therefore as follows:  -->
How large are the effects of external popularity cues such as like and dislike buttons compared to the internal weighing of pros and cons?
And can the privacy calculus be manipulated by popularity cues?
<!-- Third, in this study I approach the privacy calculus from the theoretical perspective of _bounded rationality_.  -->
<!-- It is likely that other factors next to risks and benefits also determine behavior. -->
<!-- I therefore extend the privacy calculus model theoretically by investigating the role and interplay of trust and self-efficacy. -->
<!-- To answer these research questions, a representative sample of the German population was collected in a preregistered online field experiment. -->
<!-- Participants were randomly distributed to one of three different websites, which either included a like button, both a like and a dislike button, or no buttons at all.  -->
<!-- Over the course of one week, participants discussed a topical issue (i.e., prevention of terrorist attacks in Germany).  -->
<!-- In a follow-up questionnaire, they answered items measuring the privacy calculus variables. -->

@@

The privacy calculus analyzes why people communicate online.
When are we willing to engage in a conversation?
<!-- Do they express their opinion?  -->
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977] and states that people are weighing risks and benefits before communicating.
<!-- In communicating online, people share much information about themselves. -->
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Disclosing information can be risky, and if people are more concerned about their privacy 
<!-- they should be less likely to communicate information. -->
<!-- Almost all types of communication allow for meaningful inferences about a person, even when we _don't_ communicate [@watzlawickPragmaticsHumanCommunication2011].  -->
<!-- However, the recent years have illustrated vividly how much additional insights we can get about a person, simply by analyzing their digital traces of communication or their meta-data [@kosinskiPrivateTraitsAttributes2013].  -->
<!-- The more we communicate, the more we self-disclose. -->
<!-- Online privacy concerns represent "[...] how much an individual is motivated to focus on their control over a voluntary withdrawal from other people or societal institutions on the Internet, accompanied by an uneasy feeling that their privacy might be threatened" [@dienlinLongitudinalAnalysisPrivacy2021, p. 4].  -->
<!-- Notably, I don't assume that this weighing process is flawless or perfectly rational. -->
<!-- Instead, I understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990], which has three main tenets:  -->
<!-- "(1) humans are cognitively constrained;  -->
<!-- (2) these constraints impact decision making;  -->
<!-- and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303] -->
<!-- Crucially, although bounded rationally upholds that human behavior is not perfectly logical, this does not meant that it is illogical [@gigerenzerBoundedRationalityAdaptive2002]. -->
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
<!-- Transferred to the context of online privacy, it is by now well known that several irregularities and inconsistencies between concerns and communication behavior exist. -->
<!-- These differences stem from, for example, information asymmetries, present bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020]. -->
<!-- Indeed, If people experience more concerns,  -->
they are less likely to communicate online [@baruhOnlinePrivacyConcerns2017; @dienlinExtendedPrivacyCalculus2016a; @krasnovaOnlineSocialNetworks2010; @masurUnderstandingEffectsConceptual2023; @meierPrivacyCalculusRevisited2022].
According to a meta-analysis by @baruhOnlinePrivacyConcerns2017, the effect is small but significant (*r* = -.13).
<!-- At the same time, _on average_ people do behave according to their interests, respond to incentives, or actively manage their privacy [@baruhOnlinePrivacyConcerns2017; @braunlichLinkingLooseEnds2020; @dienlinExtendedPrivacyCalculus2016a; @soloveMythPrivacyParadox2020]. -->
<!-- I therefore hypothesize that people who experience more privacy concerns engage in less communication online. -->
<!-- In light of bounded rationality and the existence of other competing factors that also influence online-communication (see below), the effect is likely small. -->

Outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016a; @bolUnderstandingEffectsPersonalization2018b; @kezerGettingPrivacyCalculus2022].
People are happy to trade in parts of their privacy when they can receive something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support [@krasnovaOnlineSocialNetworks2010], social capital [@ellisonBenefitsFacebookFriends2007], entertainment [@dhirUnderstandingRelationshipIntensity2017], information-seeking [@whitingWhyPeopleUse2013], and self-presentation [@minHowArePeople2015].
<!-- As a result, we expect a moderate relationship. -->

@@

Although privacy calculus implies that people explicitly compare benefits and drawbacks before communicating online, prior research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if and by how much people actively deliberate about their privacy, and whether doing so influences their decision to communicate.
<!-- We can understand the privacy calculus from two perspectives. -->
<!-- First, is the communication behavior aligned with people's privacy concerns and expected benefits?  -->
<!-- Second, is the communication process implicit or explicit? -->
<!-- Here, I suggest that the privacy calculus should be discussed in light of dual process theories, which state that people either deliberately, explicitly, and centrally take decisions, or instead do so automatically, implicitly, and peripherally [@kahnemanThinkingFastSlow2011; @pettyCommunicationPersuasionCentral1986].  -->
<!-- Accordingly, privacy calculus would assume that people, when it comes to disclosing, engage in a central processing.  -->
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits of communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.

How could deliberating about one's privacy affect communication?
On the one hand, it could reduce subsequent communication.
Refraining from communication---the primary means of connecting with others [@altmanPrivacyConceptualAnalysis1976]---often requires restraint [@omarzuDisclosureDecisionModel2000]. 
This is especially true for social media, which are designed to foster communication and participation [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021].
Actively thinking about whether communicating is worthwhile might be the first step not to participate.
On the other hand, deliberating about privacy might also increase communication.
In light of the benefits mentioned above, it can make much sense to conclude that participation is beneficial, thereby fostering communication [@krasnovaOnlineSocialNetworks2010]?
<!-- I therefore formulate the following research question: -->
<!-- Deliberation could represent some kind of inner consent, providing additional affirmation. -->
<!-- Alternatively, it could be that deliberation functions as a moderator. -->
<!-- For example, if people actively deliberate about whether or not to disclose, this might reinforce the effect of concerns or gratifications. -->
<!-- Reflecting about the pros and cons of communication might concerns and gratifications more salient.  -->
<!-- Alternatively, it could also be that deliberating decreases the effects, for example because apparent gratifications are considered more critically, and maybe loose their appeal. -->

@@

<!-- RQ2: Do people who deliberate more actively whether they should communicate, show larger or smaller relations between concerns, gratifications and communication behavior? -->
This weighing process is not flawless or perfect.
It is therefore useful to understand the privacy calculus from a perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
Importantly, although human behavior is _partly_ irrational, bounded rationality does not state that it is _completely_ irrational [@gigerenzerBoundedRationalityAdaptive2002].
Rationality is a continuum. 
<!-- Instead, it makes more sense to consider rationality a continuum. -->
<!-- Humans are still trying to optimize the outcomes of their behavior according to their own best interests or values. -->
<!-- It is only that their capacity to do so is bounded. -->
In the context of online privacy, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors next to gratifications and concerns should also explain communication.
Factors less focused on rationality, but more on socio-psychological aspects.
<!-- Communication online often takes place in situations where information is limited or obscure. -->

Prior research suggests that two factors are most important to understand online communication: self-efficacy and trust [correaParticipationDivideOnline2010; hossainICTEnabledVirtual2004; @metzgerPrivacyTrustDisclosure2004].
For example, if users are more familiar, experienced, and knowledgeable in a given online context, they are more likely to navigate that online contexts successfully and to communicate actively [@baruhOnlinePrivacyConcerns2017; @kramerMasteringChallengeBalancing2020].
People with more privacy self-efficacy also engage more successfully in self-withdrawal [@dienlinExtendedPrivacyCalculus2016a].
Hence, if users possess more _self-efficacy_ to participate, they should also communicate more.

@@

In all situations where people lack experience, control, or competence, a central variable to understand behavior is _trust_ [@gefenTrustTAMOnline2003].
Especially in online contexts, trust plays a key role [@metzgerPrivacyTrustDisclosure2004], as users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
<!-- Whereas specific trust focuses on the causes of trust, general trust emphasizes the experience of trust. -->
In online contexts, there are different targets of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011]. 
<!-- @gefenTrustTAMOnline2003 prioritize specific trust (p. 60).  -->
<!-- Because the targets can be largely different, it is often recommended to analyze them individually. -->

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
How easily can the calculus be affected externally?
One of the central tools to afford and govern online behavior are so-called popularity cues such as like and dislike buttons [@stsiampkouskayaNotExperimentalStudy2023].
Popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020].
Online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].

How and why might popularity cues affect the privacy calculus?
In analyzing this question, it makes sense to analyze the cues' underlying _affordances_ [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
<!-- ; as such, they are by definition subjective.  -->
There is an ongoing debate on how to best understand and operationalize affordances [@evansExplicatingAffordancesConceptual2017].
Whereas @evansExplicatingAffordancesConceptual2017 propose three affordances for mediated communication (i.e., anonymity, persistence, and visibility), @foxDistinguishingTechnologiesSocial2017 suggest 10 affordances for SNSs alone (i.e., accessibility, bandwidth, social presence, privacy, network association, personalization, persistence, editability, conversation control, and anonymity).
<!-- The privacy calculus states that both benefits and costs determine behavior. -->
Popularity cues such as like and dislike buttons are understood as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142].
<!-- Both can be linked to the two sides of the privacy calculus.  -->
<!-- The like button is positive and a potential benefit: -->
<!-- It expresses an endorsement, a compliment, a reward [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].  -->
<!-- The dislike button is negative and a potential cost:  -->
<!-- It expresses criticism and a way to downgrade content.  -->
<!-- In light of their potentially strong negative effects, most major social networking sites did not implement dislike buttons.  -->
<!-- At the time of writing, only a handful of websites have (partially) implemented it (e.g., youtube, stackexchange, or reddit). -->

Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might afford active communication. 
The mere option to receive dislikes, conversely, might inhibit communication.
Second, _actually receiving_ likes or dislikes might affect behavior, by means of positive reinforcement (likes) or punishment (dislikes) [@skinnerScienceHumanBehavior2014].
To illustrate, likes are affirmative and represent the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like-buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate. 
Dislikes, instead, represent a punishment, introducing a _loss frame_. 
<!-- Although most communication emphasizes positive aspects, the Internet is also replete with spite, envy, and arguments. -->
Websites featuring both like _and_ dislike buttons should therefore be more ambivalent compared to websites without popularity cues.
<!-- Especially in online contexts, short-term gains often outweigh long-term losses [@acquistiSecretsLikesDrive2020]. -->
<!-- Having both like and dislike buttons might still lead to more gratifications and communication. -->
Privacy concerns should not be reduced anymore:
People who are more concerned about their privacy are also more shy and risk averse [@dienlinPsychologyPrivacyAnalyzing2017].
Implementing the dislike button might therefore increase privacy concerns, thereby canceling out the positive effects of the like button. 
<!-- And because there is more at stake, participants should deliberate even more whether or not to disclose.  -->

@@

<!-- When directly comparing websites including both like and dislike buttons with websites including only like buttons, building on the rationales presented above, it is likely that websites including both buttons should increase privacy concerns and privacy deliberation. -->

@@

This manuscript features a [companion website](https://XMtRa.github.io/like_dislike), which includes the data, research materials, analyses scripts, and a reproducible version of this manuscript (see https://XMtRa.github.io/like_dislike).
The hypotheses, sample size, research materials, analyses, and exclusion criteria were preregistered (see https://osf.io/a6tzc/?view_only=5d0ef9fe5e1745878cd1b19273cdf859). 
In some cases, the preregistered approach needed to be changed (see [companion website](https://XMtRa.github.io/like_dislike/preregistration_changes)). 
Analyses that were not preregistered are reported as exploratory analyses.

@@

The data were collected in Germany.
Participants were recruited using the professional panel agency Norstat.
As incentive, participants were awarded digital points, which they could use to get special offers from online retailers. 
Participants had to be above the age of 18 and had to live in Germany. 
In a first step, the company sent its panel members an invitation to participate in the study. 
In this invitation, panel members were asked to participate in a study analyzing the current threat posed by terrorist attacks in Germany.
<!-- [^1]  -->
Members who agreed to participate were sent the first questionnaire (_T1_).
The questionnaire asked participants about their sociodemographic background, (b) provided more details about the study, and (c) included a registration link for the website, which was introduced as "participation platform".
Afterward, participants were randomly assigned to one of the three websites. 
After registration was completed, participants were invited (but not obliged) to visit the website and to discuss the topic of the terrorism threat in Germany. 
Participants could use the website and write comments over the course of one week.
Subsequently, participants received a follow-up questionnaire in which the self-reported measures reported below were collected (_T2_). 
<!-- Measures were collected after and not before the field phase in order not to prime participants or reveal the primary research interest. -->

<!-- [^1]: Although the terror attack was not of primary interest for this study, the data can and will also be used to analyze perceptions of the terrorism threat. Hence, no deception took place, and in the debriefing participants were informed about the additional research interest in privacy.  -->

@@

<!-- ### Trust -->
